Omics-based Single Step Trait Prediction

### 2017-09-06 paper/manuscript.tex
paper, describe bootstrap
paper, captions for new figures

\Genetics2level{Population structure}
Population structure in Exp2 was examined by running a STRUCTURE
\cite{Pritchard2000} analysis with the number of putative ancestral
populations $K$ ranging from 2 to 9.
Based on the estimated probability of the data for each $K$, we chose $K=4$
and ran a principal component analysis (PCA) on the scaled and centered
genomic feature matrix using the \emph{pca} function from the R-pacakge LEA
\cite{Frichot2015}.
Least-squares estimates of ancestry proportions were estimated based on sparse 
nonnegative matrix factorization algorithms~\cite{Frichot2014} using the 
\emph{snmf} function in \textit{LEA}.
Genotypes in the PCA plot (Fig.~\ref{fig:pca}) were colorized based on their
primary assignment to any of the four putative ancestral populations.
To avoid inflated predictive abilities, which can arise from large differences
in the average performance of the different subpopulations
\cite{Windhausen2012}, we reduced the panel of inbred lines to those primarily
belonging to either cluster 1 or cluster 4 in Fig.~\ref{fig:pca}.

@@

@@

@@

% Script: ./analysis/maizego_core_fraction_plot_predictive_abilities.R
\begin{figure}[H]
  \centering
  \includegraphics{./tables_figures/inbred_line_combi_plot.pdf}
  \caption{
  Predictive abilities and average coefficient of variation for
  tropical/subtropical inbred lines from the maize diversity panel and six 
  agronomic traits.
  As predictors, genomic (G) data, transcriptomic (T) data and their 
  combination (GT) were used.
  (\textbf{A}) The `Core' panel includes a subset of 149 lines covered by
  genomic and transcriptomic information whereas the `Full' panel comprises all
  211 genotypes. In the `Full' set of genotypes, transcriptomic records were
  imputed for 62 lines.
  (\textbf{B}) Influence of structural missingness in the coverage of genotypes 
  with transcriptomic information on single step-based predictive abilities 
  using a `Core' set of 149 inbred lines.
  Transcriptomic records were deleted for 17 to 19 genotypes that were
  primarily associated with a particular subgroup in the material, as defined
  by an admixture analysis and a PCA (Fig.~\ref{fig:PopStructure}).
  (\textbf{C}) Influence of core set sampling on predictive abilities using
  nine different sub set sizes where `Core Fraction' indicates the fraction
  of 149 inbred lines that is covered by both, genomic and transcriptomic
  data.
  }
\label{fig:InbredResults}
\end{figure}

@@

% Script: ./analysis/uhoh_plot_predictive_abilities.R
\begin{figure}[H]
\centering
  \includegraphics{./tables_figures/hybrid_predictive_ability.pdf}
  \caption{
    Predictive abilities and average coefficients of variation for the set of 
    maize hybrids and six agronomic traits.
    As predictors, pedigree (P), genomic (G), transcriptomic (T) data and 
    combinations thereof we used.
    The `Full' set of genotypes comprises 1,521 hybrid progeny based on 245
    parental inbred lines, whereas the `Core' set comprises 685 hybrid progeny
    based on 103 parental inbred lines.
  }
\label{fig:HybridResults}
\end{figure}

\Genetics2level{Coverage of the genetic space}
Single-step approaches are based on the use of one complete predictor that
covers all genotypes and one incomplete predictor that covers only a subset of
all available genotypes.
Here, we explore the impact of which data are covered by the incomplete 
predictor on the predictive abilities obtained via single-step approaches by
focusing exclusively on the 149 inbred lines from the maize diversity panel
that are covered by both, genomic and transcriptomic information.
First, we assembled nine different core sets, which we defined as subsets
of genotypes that maximize the average genetic distance between genotypes of
the full set.
The nine core sets differ in that each covers a different number of genotypes,
ranging from only 10\% to 90\% of all genotypes covered by the incomplete
predictor (Fig.~\ref{fig:CoreSetPCA}).

\begin{figure}[H]
\centering
\includegraphics{./tables_figures/maizego_core_sampling_pca.pdf}
\caption{
  Coverage of the genetic space by nine different core sets. 
  Genotypes are distributed according to their scores based on the first two
  principal components (PC).
  The total genetic space was spanned by 149 genotypes from the maize diversity 
  panel for which both, genomic and transcriptomic data, were available.
  Genotypes that were members of the core set (orange triangles) were declared 
  as being covered by both predictors whereas members of the complement (blue,
  filled circles) were declared as only being covered by genomic information.
  Assignments of genotypes to the core set was done by selecting a subset of
  genotypes that maximized the average genetic distance in the core set based 
  on the Modified Rogers distance between all genotypes.
  Core set sizes ranged from 10\% to 90\% of the 149 available genotypes and
  are given as decimals above each facet of the plot.
}
\label{fig:CoreSetPCA}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics{./tables_figures/maizego_admixture_k3_pca.pdf}
\caption{
  Population structure of 149 maize inbred lines from the maize diversity panel
  covered by genomic and transcriptomic data.
  (\textbf{A}) Ancestry coefficients from an admixture analysis when 
  considering $K=3$ ancestral populations.
  (\textbf{B}) Principal component analysis with genotypes colored according to 
  the most likely ancestral population (\textit{i.e.}, share $\geq 0.5$) as 
  indicated by the admixture analysis.
  Genotypes for which none of the three ancestry coefficients reached a value
  of 0.5 were classified as ``mixed''.
}
\label{fig:PopStructure}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics{./tables_figures/diversity_dent_flint_pca.pdf}
\caption{
  Principal component (PC) analysis i) of the 211 tropical/subtropical lines 
  from the maize diversity panel, ii) of the 142 Dent parent lines of the 1,521
  maize hybrids and iii) of the 103 Flint parent lines of the maize hybrids.
  All inbred lines, which are covered by genomic and transcriptomic features
  are depicted as orange triangles whereas genotypes for which only genomic
  information are available are depicted as blue dots.
}
\label{fig:PCA}
\end{figure}

>>>>

\Genetics2level{Population structure}
Population structure in Exp2 was examined by running a STRUCTURE
\cite{Pritchard2000} analysis with the number of putative ancestral
populations $K$ ranging from 2 to 9.
Based on the estimated probability of the data for each $K$, we chose $K=4$
and ran a principal component analysis (PCA) on the scaled and centered
genomic feature matrix using the \emph{pca} function from the R-pacakge LEA
\cite{Frichot2015}.
Least-squares estimates of ancestry proportions were estimated based on sparse 
nonnegative matrix factorization algorithms~\cite{Frichot2014} using the 
\emph{snmf} function in \textit{LEA}.
Genotypes in the PCA plot (Fig.~\ref{fig:inbred-pca}) were colorized based on their
primary assignment to any of the four putative ancestral populations.
To avoid inflated predictive abilities, which can arise from large differences
in the average performance of the different subpopulations
\cite{Windhausen2012}, we reduced the panel of inbred lines to those primarily
belonging to either cluster 1 or cluster 4 in Fig.~\ref{fig:inbred-pca}.

\begin{figure}[H]
  \centering
  \includegraphics{./tables_figures/inbred_pca.pdf}
  \caption{
  Principal component analysis of 211 genotypes from the set
  `tropical/subtropical' of the maize diversity panel inbred lines.
  The four clusters are based on $K=4$ putative ancestral populations in a
  STRUCTURE analysis of the corresponding genomic feature matrix.
  Genotypes are colored based on their primary assignment to any of these four
  clusters.
}
\label{fig:inbred-pca}
\end{figure}

@@

Standard errors for the predictive abilities were calculated by bootstrapping
predictive and observed values 1,000 times using the \emph{R} package 
\emph{boot}~\cite{Canty2017}.

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=1\textwidth]{./tables_figures/compound_plot.pdf}
  \caption{
  (\textbf{a}) Scheme for the amount of genotypes covered by predictors.
  The two extended bars represent the entire set of genotypes in either Exp1 or
  Exp2.
  All genotypes are covered by at least one predictor, which is called the
  `complete' predictor and depicted as a white bar.
  The second bar depicts that only a fraction of genotypes is covered by both,
  the `complete' and the `incomplete' predictor (dark gray) and that the
  difference between the predictors will be imputed.
  The four bars to the right outline the principle of generating core sets for
  Exp2.
  For nine different values of $x$, which represents the fraction of genotypes
  covered by the `incomplete' predictor, core sets are build.
  (\textbf{b}) Subsampling within core sets.
  Each of the black symbols represents a different genotype.
  For each value of $x$, 100 subsets $s$ are generated where $x\%$ of genotypes
  are randomly declared to be covered by both, the `complete' and the
  `incomplete' predictor whereas the remaining $100\% - x\%$ of genotypes
  declared to be covered only by the `complete' predictor.
  }
\label{fig:compound-plot}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics{./tables_figures/pred_ability_inbred.pdf}
  \caption{
  Predictive abilities and SE for tropical/subtropical inbred lines from the 
  maize diversity panel and six agronomic traits.
  As predictors, genomic (G) data, transcriptomic (T) data and their 
  combination (GT) were used.
  The `Core' panel includes a subset of 110 lines covered by genomic and
  transcriptomic information whereas the `Full' panel comprises all 164
  genotypes.
  In the `Full' set of genotypes, transcriptomic records were imputed for 54
  lines.
  }
\label{fig:InbredResults}
\end{figure}

@@

\begin{figure}[H]
\centering
  \includegraphics{./tables_figures/pred_ability_hybrid.pdf}
  \caption{
    Predictive abilities and average, bootstrapped standard errors for the set
    of maize hybrids and six agronomic traits.
    As predictors, pedigree (P), genomic (G), transcriptomic (T) data and 
    combinations thereof we used.
    The `Full' set of genotypes comprises 1,521 hybrid progeny based on 245
    parental inbred lines, whereas the `Core' set comprises 685 hybrid progeny
    based on 103 parental inbred lines.
  }
\label{fig:HybridResults}
\end{figure}

\Genetics2level{Coverage of the genetic space}
Single-step approaches are based on the use of one complete predictor that
covers all genotypes and one incomplete predictor that covers only a subset of
all available genotypes.
Here, we explore the impact of which data are covered by the incomplete 
predictor on the predictive abilities obtained via single-step approaches by
focusing exclusively on the 149 inbred lines from the maize diversity panel
that are covered by both, genomic and transcriptomic information.
First, we assembled nine different core sets, which we defined as subsets
of genotypes that maximize the average genetic distance between genotypes of
the full set.
The nine core sets differ in that each covers a different number of genotypes,
ranging from only 10\% to 90\% of all genotypes covered by the incomplete
predictor (Fig.~\ref{fig:inbred-pca}).

\begin{figure}[H]
  \centering
  \includegraphics{./tables_figures/pred_ability_inbred_core_fraction.pdf}
  \caption{
  Influence of the fraction of genotypes covered by the incomplete predictor
  on predictive abilities.
  Predictive abilities are based on transcriptomic effects derived from a
  single step prediction model involving transcriptomic (T) and genomic (G)
  data, respectively.
  Results are shown for the core set of 110 inbred lines from Exp2 for which
  both, G and T are available for any genotype and six agronomic traits.
  The fraction of genotypes (\textit{i.e.} `Core Fraction') for which
  T was available is given on the x-axis.
  Dotted lines indicate that $T > G$, dashed lines indicate that $T \approx G$
  and solid lines indicate that $G > T$ when running predictions using single
  predictors on the same data set.
  }
\label{fig:InbredResults}
\end{figure}

@@

@@
